Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method

### 2023-08-31 manu-draft.qmd
update sensitivity figs and text, fix #4

The prior distributions for the EBASE parameters follow truncated normal distributions with defined means ($\mu$) and standard deviations ($\sigma$) (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the results. Low and high values for the means and standard deviations for $R$ were assessed following reasonable ranges from @caffrey2004 using logic similar to that for choosing the default priors. The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, which we rounded to $1/2$ and 3 times the mean (150 and 900 $\text{mmol}/\text{m}^2/\text{d}$) and proportionally for $a$ as 2.0 and 12 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default mean values to test a range from high precision to completely uninformed values. The mean and standard deviation for the $b$ parameter were not tested and set as the default values for all analyses given that there is minimal prior knowledge on expectations for this parameter. The default prior distribution is already uninformed and we have no appropriate expectations on methods for varying this parameter. Each unique combination of low and high values for the means and standard deviations of each parameter were evaluated, creating 2$^4$ = 16 different combinations of prior distributions that were used with the synthetic time series in EBASE.

The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $R$, and $b$). Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being one day and a maximum being the length of the time series. For a chosen optimization period, single estimates for $a$, $R$, and $b$ are returned by the model. A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if variation occurs at time scales less than the optimization period. Results from optimization periods of one, seven, and thirty days were compared for the synthetic one-year time series. Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times three optimization periods) unique analyses for the assessment of changing priors of all parameters.

@@

Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values. Overall, most models regardless of optimization period or priors were able to reproduce the dissolved oxygen time series. The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $R$ parameter was low. However, substantial variation in NSE values was observed among all other parameters. Not considering dissolved oxygen, gas exchange ($D$) was estimated with the highest NSE, with a median value across all optimization periods and priors slightly larger than zero (median NSE = `r prmall[['D']]`). Conversely, the $a$ parameter was reproduced the most poorly, with a median value across all optimization periods and priors less than zero (median NSE = `r prmall[['a']]`). By optimization period, median NSE values across all parameters decreased slightly as the number of days used for model optimization increased (1 day = `r optnse[['1']]`, 7 days = `r optnse[['7']]`, 30 days = `r optnse[['30']]`). The top three ranked models were identical for the seven and thirty day models and similar for models using the one day optimization period. Median NSE values across parameters for the top model in each optimization period were also similar (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`). For the seven day and thirty day optimization periods, the NSE values for all parameters were greater than zero for the top models, where two of the parameters in the top model for the one day optimization period has NSE \< 0. Most parameters regardless of optimization period or priors had NSE \< 0 (`r perneg`). As such, the prior combinations that produced the highest NSE values across all parameters were those where the $\sigma$ values were high for each prior distribution, regardless of the values for $\mu$, i.e., those with sufficient flexibility to search a larger parameter space regardless of $\mu$ produced metabolic parameter estimates most similar to the known values from the synthetic time series.

@fig-priorsumcomp provides an alternative summary of how the prior distributions influenced the ability of EBASE to reproduce the synthetic parameters. Median NSE values across all models, parameters, and each optimization period for the different subsets of the prior combinations are shown relative to the grand median NSE value across all comparisons. For example, the bottom left panel with the x-axis value for the $a$ parameter shows the difference in the median NSE values for all models using a one-day optimization period where the $\sigma$ value for the $a$ prior distribution was low or high. The size of the points indicate the inter-quartile range of NSE values for each subset. As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\sigma$ values for each parameter. Positive values indicate relative performance greater than the overall median NSE values and negative values indicate relative performance worse than the overall median. Dissolved oxygen was generally modeled with high accuracy for all priors and it was excluded to not skew the assessment. The patterns suggest that EBASE results are more sensitive to changes in the $\sigma$ parameter as shown by the larger spread in the median results compared to those for $\mu$. This result is especially true for the priors of the $a$ and $R$ parameters. Conversely, model results are less sensitive to low or high values for $\mu$ as the spread between median NSE values is less than those for $\sigma$. This suggests that EBASE model sensitivity is most influenced by $\sigma$ of the prior distributions, i.e., larger values produce more accurate and less variable estimates regardless of $\mu$. Patterns by optimization period were not apparent.

The actual model results for the best and worst performing combinations of prior distributions for the one day, seven day, and thirty day optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability). The best performing models were those with the highest median NSE values across all parameter estimates in @fig-priorcomp. The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the parameter estimates were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the one day optimization period. However, the $b$ parameter was estimated with high variability for the one and seven day optimization periods (@fig-optex a, c, bottom row) and with a large positive bias for the 30 day optimization period (@fig-optex e, bottom row). These results were not unexpected because it was impossible to evaluate NSE estimates using a fixed value in the synthetic results and the accuracy was not considered in identifying the best performing model. Finally, the worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values). The worst performing models were caused by constraining EBASE to a small parameter space using low or high values for $\mu$ combined with low values for $\sigma$.

As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of instrument and tidal noise to the original synthetic time series. The objective of the analysis is to develop an expectation of how metabolic estimates may change using in situ data with noise that may substantially mask a biological signal in dissolved oxygen measurements. The instrument and tidal noise were estimated using a weighted regression analysis [@Beck15; @Beck21] on actual observed dissolved oxygen time series for Apalachicola Bay using data from 2021. This approach models dissolved oxygen as a function of time, salinity, and season using a moving window approach with weights of the independent variables appropriate for each time period within the windows. The results of this analysis return a predicted and detided dissolved oxygen time series. The instrument noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series. Both the instrument and tidal noise were added to the original synthetic time series above (@fig-synapanos). EBASE was then used with unconstrained priors (large $\sigma$) on the synthetic time series with noise, then compared with results from the original synthetic time series.

@@

![Comparison of EBASE results to a synthetic time series created with known model parameters. ($a$, $R$, and $b$). Results are compared for unique combinations of prior distributions and optimization period. The means ($\mu$) and standard deviations ($\sigma$) are evaluated as low (L) or high (H) values (see text for details) and the optimization period for EBASE is one day, seven days, or thirty days. The model predictions of dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values. NSE values from zero to one (green) indicate EBASE has some skill in recovering the synthetic values of DO, $D$, $P$, $R$ , and $a$, with values closer to 1 being a better approximation, whereas negative values indicate no skill. The color scale for NSE is also exponentially transformed to minimize the influence of large negative values on the visual display. The $b$ parameter is a fixed constant and was not evaluated. Rankings of each combination as the median NSE value across all five model outputs in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}

@@

![Summaries of % change in Nash-Sutcliffe Efficiency (NSE) values across all the comparisons of EBASE and the five synthetic model outputs separated by different combinations of prior values. Summaries are shown relative to the change from the median for each comparison group (i.e., optimization period and prior parameter, $n$ = 32). Each result on the x-axis represents the median values for which the combination prior applies. For example, the points in the upper left panel that are fartheset to the left summarize the results from @fig-priorcomp where $\mu$ is high or low for the $a$ parameter using a one day optimization period. Point size represents the inter-quartile range of median NSE values for each result ($n$ = 8) .](figs/priorsumcomp.png){#fig-priorsumcomp}

@@

![Comparison of EBASE results for the best and worst performing models to a synthetic time series created with known metabolic parameters. The best and worst performing models for the one day (a, b), seven day (c, d), and thirty day (e, f) optimization periods are based on the EBASE results with prior distributions that produced the highest and lowest median Nash-Sutcliffe Efficency values across all parameter estimates (@fig-priorcomp). The results are averaged at the time step of the model optimization period. The $b$ parameter is fixed at 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$.](figs/optex.png){#fig-optex fig-align="center"}

>>>>

The prior distributions for the EBASE parameters follow truncated normal distributions with defined means ($\mu$) and standard deviations ($\sigma$) (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]). The sensitivity analysis evaluated changes in the distributions by varying the means and standard deviations from small to large to test the effect of changes in the central tendencies and more or less constrained ranges, respectively, on the results. Low and high values for the means and standard deviations for $R$ were assessed following reasonable ranges from @caffrey2004 using logic similar to that for choosing the default priors. The minimum and maximum mean values for $R$ were evaluated as 138 and 1009 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$, which we rounded to $1/2$ and 3 times the mean (150 and 900 $\text{mmol}/\text{m}^2/\text{d}$) and proportionally for $a$ as 2.0 and 12 $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$. The minimum and maximum values for the standard deviations for $R$ and $a$ were chosen as 10% and 1000% of the default mean values to test a range from high precision to completely uninformed values. The mean and standard deviation for the $b$ parameter were not tested and the prior value was set as the fixed constant used in the synthetic time series (0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$). There is minimal prior knowledge on expectations for the $b$ parameter and initial assessments with the synthetic time series demonstrated that the parameter was poorly constrained, primarily because the fixed parameter could not be estimated using observed data (i.e., wind, water temperature) that affected the actual realization of $b$. As such, each unique combination of low and high values for the means and standard deviations for $a$ and $R$ were evaluated, creating 2$^4$ = 16 different combinations of prior distributions that were used with the synthetic time series in EBASE.

The optimization period used for a time series determines how much data are used to estimate the $a$ and $R$ parameters. Unlike BASE, the EBASE software allows flexibility in the period of time used for optimization, with a minimum being one day and a maximum being the length of the time series. For a chosen optimization period, single estimates for $a$ and $R$ are returned by the model ($b$ is also returned, but was fixed for the sensitivity analysis). A short period of optimization (e.g., one day) may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE. As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation, although at the risk of overgeneralizing metabolic rates if variation occurs at time scales less than the optimization period. Results from optimization periods of one, seven, and thirty days were compared for the synthetic one-year time series. Each optimization period was also evaluated for every unique combination of prior distributions, totaling 48 (16 times three optimization periods) unique analyses for the assessment of changing priors of all parameters.

@@

Several conclusions can be made from the results in @fig-priorcomp following the patterns of the NSE values. Overall, most models regardless of optimization period or priors were able to reproduce the dissolved oxygen time series. The only models where the NSE values were less than zero for dissolved oxygen were those where $\sigma$ of the $a$ and $R$ parameters was low and $mu$ for one of the $a$ and $R$ parameters was high. However, substantial variation in NSE values was observed among all other parameters. Not considering dissolved oxygen, gas exchange ($D$) was estimated with the highest NSE, with a median value across all optimization periods and priors slightly larger than zero (median NSE = `r prmall[['D']]`). Conversely, the $a$ parameter was reproduced the most poorly, with a median value less than zero across all optimization periods and priors (median NSE = `r prmall[['a']]`). By optimization period, median NSE values across all parameters increased as the number of days used for model optimization increased (1 day = `r optnse[['1']]`, 7 days = `r optnse[['7']]`, 30 days = `r optnse[['30']]`). The top three ranked models were identical across optimization periods and median NSE values across parameters for the top models increased with optimization period (1 day = `r optmod[['1']]`, 7 days = `r optmod[['7']]`, 30 days = `r optmod[['30']]`). The NSE values for all parameters were greater than zero for the top three models in the 30 day optimization period, whereas the $a$ parameter had NSE less than zero for the top model in the 7 day optimization period and $R$ and $a$ had negative NSE values for the top model in the one day optimization period. About half of the parameters regardless of optimization period or priors had NSE \< 0 (`r perneg`). As such, the prior combinations that produced the highest NSE values across all parameters were those where all prior distributions had low $\mu$ and $\sigma$ values for both $a$ and $R$, although the second and third best performing models had high $\sigma$ for $R$.

@fig-priorsumcomp provides an alternative summary of how the prior distributions influenced the ability of EBASE to reproduce the synthetic parameters. Median NSE values across all models, parameters, and each optimization period for the different subsets of the prior combinations are shown. For example, the left-most bars in top left panel with the x-axis value for the $a$ parameter shows the median NSE values for all models using a one-day optimization period where the $\mu$ value for the $a$ prior distribution was low or high. The lines indicate the inter-quartile range of NSE values for each subset. As such, each set of points denoted by the x-axis parameters in each panel include median NSE values for all models by optimization period, but subset by the low/high $\mu$ or $\sigma$ values for each parameter. Dissolved oxygen was generally modeled with high accuracy for all priors and it was excluded to not skew the assessment. The patterns suggest that EBASE results are more sensitive to model optimization period as shown by higher median NSE values with longer optimization times (note the different y-axis scales, also confirmed in @fig-priorplot). Similarly, median values were much more variable for the one day optimization period, as shown by a mix of both positive and negative values in the interquartile ranges. Lower median NSE values were observed for higher values for $\mu$, where higher median values for were observed for higher values of $\sigma$, regardless of optimization period.

The actual model results for the best and worst performing combinations of prior distributions for the one day, seven day, and thirty day optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for visual comparability). The best performing models were those with the highest median NSE values across all parameter estimates in @fig-priorcomp. The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the parameter estimates were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the one day optimization period. The worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values). The worst performing models were caused by constraining EBASE to a small parameter space, such as a high $\mu$ and low $\sigma$ for $a$ as in the bottom row for subplot b, d, and f.

As a final evaluation of EBASE to return known metabolic parameters, a second synthetic time series was created that added estimated effects of noise to the original synthetic time series. The objective of the analysis is to develop an expectation of how metabolic estimates may change using in situ data with noise that may substantially mask a biological signal in dissolved oxygen measurements. The noise estimates were created using a weighted regression analysis [@Beck15; @Beck21] on actual observed dissolved oxygen time series for Apalachicola Bay using data from 2021. This approach models dissolved oxygen as a function of time, salinity, and season using a moving window approach with weights of the independent variables appropriate for each time period within the windows. The results of this analysis return a predicted and detided dissolved oxygen time series. The instrument noise was estimated as the difference between the observed and predicted time series and the tidal noise was estimated as the difference between the predicted and detided time series. Both the instrument and tidal noise were added to the original synthetic time series above (@fig-synapanos). EBASE was then used with unconstrained priors (large $\sigma$) on the synthetic time series with noise, then compared with results from the original synthetic time series.

@@

![Comparison of EBASE results to a synthetic time series created with known model parameters. ($a$, $R$, and $b$). Results are compared for unique combinations of prior distributions and optimization period. The means ($\mu$) and standard deviations ($\sigma$) are evaluated as low (L) or high (H) values (see text for details) and the optimization period for EBASE is one day, seven days, or thirty days. The model predictions of dissolved oxygen (DO), gross primary production (P), ecosystem respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on Nash-Sutcliffe Efficiency (NSE) values between the EBASE results and synthetic values. The $b$ parameter is a fixed constant and was not evaluated. Rankings of each combination as the median NSE value across all five model outputs in a row are shown on the right, with the top three as black/bold and bottom three as black/italic.](figs/priorcomp.png){#fig-priorcomp}

@@

![Summaries of Nash-Sutcliffe Efficiency (NSE) values across all the comparisons of EBASE and the five synthetic model outputs separated by different combinations of prior values (note the different y-axis scales). Each result on the x-axis represents the median values for which the combination prior applies. For example, the bars in the upper left panel that are farthest to the left summarize the results from @fig-priorcomp where $\mu$ is high or low for the $a$ parameter using a one day optimization period. Bars are the medians and lines the interquartile ranges of NSE values for each result ($n$ = 8 for each bar) .](figs/priorsumcomp.png){#fig-priorsumcomp}

@@

![Comparison of EBASE results for the best and worst performing models to a synthetic time series created with known metabolic parameters. The best and worst performing models for the one day (a, b), seven day (c, d), and thirty day (e, f) optimization periods are based on the EBASE results with prior distributions that produced the highest and lowest median Nash-Sutcliffe Efficency values across all parameter estimates (@fig-priorcomp). The results are averaged at the time step of the model optimization period.](figs/optex.png){#fig-optex fig-align="center"}
